\chapter{Conclusion}
\label{sec:conclusion}
% Conclusion/Summary

This project focused on evolving and extending formal certificate generation for a program verifier.
Building upon the foundational work of Parthasarathy et al.,
which established a semantic formalization and certificate generator for a core subset of the Boogie language in Isabelle/HOL.
We first evolved the architecture by updating the underlying Boogie codebase to the latest version, resolving related compatibility issues, and ensuring the tool can be actively run and developed for future contributions.
Following a systematic analysis to identify the most critical unsupported features, we selected Boogie maps as our primary target for extension.

Consequently, we designed and implemented formal semantics for monomorphic, single-arity Boogie maps with finite domain nesting.
We achieved this through a shallow embedding into Isabelle/HOL functions governed by custom well-formedness constraints.
This approach successfully demonstrates the practical possibilities
and inherent limitations of shallow embeddings when contrasted
with deeper embeddings that use expressions and explicit variable binders.

Our formalized map semantics guarantee extensionality while requiring only minimal modifications to the predicate verification condition axioms.
Furthermore, by augmenting the proof generation module with rudimentary support for maps,
we enabled the first successfully verifiable certificate generation for map-dependent Boogie programs, validating the soundness of our approach.
While this establishes a robust foundation, extending the proof tactics to automate a broader range of certificate verifications,
and generalizing the semantics to natively support complex map behaviors,
such as lambda expressions, arbitrary domain nesting, and polymorphic maps, remain open challenges for future work.